---
title: "Simple Tree Building in R"
author: "Samantha Lycett"
date: "23 Feb 2024"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
library(tibble)
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

figPath <- paste0("simple_tree_building_in_R/")
dir.create(figPath)

knitr::opts_chunk$set(fig.path = figPath)
knitr::opts_chunk$set(fig.width=8, fig.height=8)
```

# Simple Tree Building in R

## Automating Trees in R

R package ape provides some functions which enable reading sequences from fasta files; Neighbour joining tree building; Simple tree plotting; Finding the best root and performing a root-to-tip plot; from Root-to-Tip plot you can find the estimated clock rate and time to most recent common ancestor.

Here is example R script to show how a Neighbour joining tree can be built from a set of fasta sequences, and how to perform a Root-to-Tip molecular clock analyses (similar to Tempest).

```{r}
# install R package ape if needed (only do this once)
#install.packages("ape")

# load required library (do this every time)
library(ape)

# load custom functions (utility code for sequence or tree tips names)
source("getEl.R")
source("calcDecimalDate.R")
```

## STEP 1 - read sequences from file

```{r}
# read sequences into seqs object using R package ape and read.dna function
seqFileName <-  "H5N1_HA_sel5regions.fas"
seqs        <- read.dna(seqFileName,format="fasta",as.matrix=FALSE)

# number of sequences
nseqs       <- length(seqs)
```

There were `r nseqs` sequences read from `r seqFileName`

The sequence names are within the seqs object, the first 5 are:
```{r}
attributes(seqs)$names[1:5]
```
Here we can see the format of the sequences names, it is Accession|Host-type|Region|IsolateName|decimalDate

The distribution of Host-types in the data set is:
```{r}
taxa <- as.matrix(attributes(seqs)$names)
host <- apply(taxa,1,getEl,ind=2,sep="\\|")
kable(t(table(host)), caption="Host-types")
```

```{r}
region <- apply(taxa,1,getEl,ind=3,sep="\\|")
kable(t(table(region)), caption="Region")
```

## STEP 2 - make distance matrix

Use the TN93 (Tamura-Nei 93) nucleotide substitution model, assume gamma distributed rates between sites with a gamma distribution shape parameter of 0.1, and allow pairwise deletion

```{r}
dd          <- dist.dna(seqs,model="TN93",gamma=0.1,pairwise.deletion=TRUE)
```

## STEP 3 - make Neighbour joining tree

Now use the just created distance matrix to make a neighbour joining tree with function nj.  Plot the tree without showing the tip labels

```{r, simple_tree_raw_tree}
tr  <- nj(dd)
plot(tr, show.tip.label=FALSE)
```

Reorder the branches of the tree using function ladderize and replot the tree (notice how it looks different)

```{r, simple_tree_ladderized_tree}
tr  <- ladderize(tr)
plot(tr, show.tip.label=FALSE)
```

Save the tree to file in newick format

```{r}
treeFileName<- gsub(".fas","_ape_tn93_nj.nwk",seqFileName)
write.tree(tr, file=treeFileName)
```

The tree has been written to file `r treeFileName`

## STEP 4 - Rooting the tree using root-to-tip

Extract the decimal dates from the carefully constructed sequence names.  The sequence names have the decimal date encoded at the end separated by a bar.

The first sequence name in the tree (first taxa) is: `r tr$tip.label[1]`

The first 5 sequences in the tree are the following (note the order is not necessarily the same as in the sequences file used to generate the tree):

```{r}
tr$tip.label[1:5]
```


```{r}
# uses custom function getEl
decDates  <- as.numeric(apply(as.matrix(tr$tip.label),1,getEl,ind=1,fromEnd=TRUE,sep="\\|"))
```

Use the rtt (root to tip) function from R package ape, this uses the decimal dates and tries to find the best fitting root.  It returns a rooted tree.

```{r}
# find best rooted tree using rtt function and decimal dates (decDates)
rtr <- rtt(tr,decDates)

# save the best rooted tree
tname<- gsub(".nwk","_rtt.nwk",treeFileName)
write.tree(rtr, file=tname)
```

```{r, rooted_nj_tree}
plot(rtr, show.tip.label=FALSE)
title("Rooted NJ tree")
add.scale.bar()
```

## Step 5 - Get Root to tip plot and extract clock rate and TMRCA

Get root to tip information, this is the genetic distance from the root.  Use function dist.nodes - which calculated the genetic distance from the root node to all the nodes of the tree.

```{r}
# get the genetic distance (all nodes)
rtt_all_dists<- dist.nodes(rtr)
ntips        <- length(rtr$tip.label)
rootNode     <- length(rtr$tip.label)+1

# get the genetic distance from the rootNode to all the sequences (tips)
rootToTip    <- rtt_all_dists[1:ntips,rootNode]
```

Fit a straight line (i.e. linear model, lm) to the genetic distance from root (rootToTip) vs time (decDates)

```{r}
clock_fit    <- lm(rootToTip ~ decDates)
```

Extract the slope of the line from the line fit - which is the clock rate,  but round it abit for easier printing to screen.

```{r}
clockrate    <- round(clock_fit$coefficients[2]*100000)/100000
```

Extract the intercept of the line - this is the TMRCA, but round it abit for easier printing to screen

```{r}
tmrca        <- -clock_fit$coefficients[1]/clock_fit$coefficients[2]
tmrca        <- round(tmrca*1000)/1000
```

Show root to tip plot
```{r, root_to_tip_plot}
plot(decDates,rootToTip,pch=21,
     xlab="Date", ylab="Genetic Distance From Root")
abline(clock_fit,col="blue")
legend("topleft",c(paste("TMRCA =",tmrca),
                   paste("Clock rate estimate =",clockrate),
                   "Sequences"),
       pch=c(NA,NA,21), col=c(NA,"blue","black"), lty=c(NA,1,NA),bty="n")
title("Root-to-Tip plot of NJ tree")
```





